---
title: "NBA 4920/6921 Lecture 16"
subtitle: "Tree Methods: Regression Application" 
author: "Murat Unal"
date: "10/26/2021"
output: 
  beamer_presentation:
    colortheme: beaver
    df_print: kable
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = '', 
                      warning = FALSE, 
                      message = FALSE)
```

------------------------------------------------------------------------

```{r message=FALSE, warning=FALSE}
rm(list=ls())
options(digits = 3, scipen = 999)
library(tidyverse)
library(ISLR)
library(cowplot)
library(stargazer)
library(lmtest)
library(sandwich)
library(MASS)
library(jtools)
library(caret)
library(rpart)
library(rpart.plot)
library(ROCR)
set.seed(2)
```

***
```{r}
Hitters <- ISLR::Hitters
Hitters <- na.omit(Hitters)
```
- Seperate train and test data
```{r}
train = sample(1:nrow(Hitters), 0.7*nrow(Hitters))

```

***
# Regression Trees

- We can fit a regression tree using `rpart` and then visualize it using `rpart.plot`

- we need to set `method = "anova"`, for classification set `method="class"`

- The main tuning parameter is `cp`, the complexity parameter


***
```{r}
hit_tree = rpart(Salary ~ .,data = Hitters[train,],
                 method="anova")
```

***
```{r}
hit_tree
```

***
- We can visualize our tree model with `rpart.plot()`

```{r}
rpart.plot(hit_tree)
```

***
- Behind the scenes `rpart()` is automatically applying a range of cost complexity $\alpha$  values to prune the tree. 

- To compare the error for each  $\alpha$  value, it performs a 10-fold CV (by default)

```{r}
hit_tree$cptable
```

- Here we don't find much improvement after 2 terminal nodes 


***

- Notice the dashed line which goes through the point  $|T|=2$.

- It’s common to instead use the smallest tree within 1 standard error (SE) of the minimum CV error (this is called the 1-SE rule). 

- Thus, we could use a tree with just 2 terminal nodes and reasonably expect to experience similar results within a small margin of error.


```{r}
plotcp(hit_tree)
```


***

- To illustrate the point of selecting a tree with 8 terminal nodes (or 2 if you go by the 1-SE rule), we can force `rpart()` to generate a full tree by setting `cp = 0` (no penalty results in a fully grown tree).

```{r}
hit_tree2 = rpart(Salary ~ .,data = Hitters[train,],
          method="anova",control = list(cp=0,xval=10))
```

***
```{r}
plotcp(hit_tree2)
abline(v = 8, lty = "dashed")
```

***

- Make predictions on the test data with best cp
```{r}
pred.tree <- predict(hit_tree, Hitters[-train,])

rmse.min.cp <- sqrt(mean((Hitters[-train,"Salary"]-
                          pred.tree)^2))

rmse.min.cp
```
***
- `maxdepth`: is the maximum number of internal nodes between the root node and the terminal nodes.
- We could obtain a tree with 2 terminal nodes by setting `maxdepth=1`.

```{r}

hit_tree1se <- rpart(Salary ~ .,data = Hitters[train,],
          method="anova",maxdepth=1)

hit_tree1se
```

***
- Predictions and rmse
```{r}
pred.tree1se <- predict(hit_tree1se, Hitters[-train,])
rmse.1se.cp <- sqrt(mean((Hitters[-train,"Salary"]-
                            pred.tree1se)^2))
rmse.1se.cp
```

# Exercise

- Train a regression tree to  predict the crime rates (`crim`) in the Boston dataset.

```{r}
data_test <- read.csv("boston_test.csv")
data_train <- read.csv("boston_train.csv")
```

***
- Obtain the tree

```{r}
bos_tree = rpart(crim ~ .,data = data_train,
                 method="anova")

bos_tree
```

***
- Visualize the tree
```{r}
rpart.plot(bos_tree)
```

***
- Examine the cross validation errors against `cp`

```{r}
bos_tree$cptable
```

***
- Plot `cp`
```{r}
plotcp(bos_tree)
```

***
- Obtain the 1se tree

```{r}
bos_tree1se = rpart(crim ~ .,data = data_train,
                 method="anova", maxdepth=1)
bos_tree1se
```


***
- Predictions and rmse for min cp and 1se cp
```{r}
pred.bos.tree <- predict(bos_tree, data_test)
rmse.min.cp <- sqrt(mean((data_test$crim-
                            pred.bos.tree)^2))
rmse.min.cp

pred.bos.tree1se <- predict(bos_tree1se, data_test)
rmse.1se.cp <- sqrt(mean((data_test$crim-
                            pred.bos.tree1se)^2))
rmse.1se.cp
```

