---
title: "NBA 4920/6921 Lecture 17"
subtitle: "Tree Methods: Classification Application" 
author: "Murat Unal"
date: "10/28/2021"
output: 
  beamer_presentation:
    colortheme: beaver
    df_print: kable
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = '', 
                      warning = FALSE, 
                      message = FALSE)
```

------------------------------------------------------------------------

```{r message=FALSE, warning=FALSE}
rm(list=ls())
options(digits = 3, scipen = 999)
library(tidyverse)
library(ISLR)
library(lmtest)
library(sandwich)
library(jtools)
library(caret)
library(leaps)
library(future.apply)
library(glmnet)
library(rpart)
library(rpart.plot)
library(ROCR)
set.seed(2)
```

***
- We'll use the `Carseats` dataset from ISLR

```{r}
Carseats <- ISLR::Carseats
Carseats = na.omit(Carseats)
```

```{r}
dim(Carseats)
names(Carseats)
```

***
- Let’s modify the response from its original numeric variable to a categorical variable with two levels: `high` and `low`

```{r}
Carseats$Sales = as.factor(ifelse(Carseats$Sales <= 8,
                                  "Low", "High"))
```

```{r}
train = sample(1:nrow(Carseats), 0.7*nrow(Carseats))
```

***

- Grow a classification tree

```{r}
sales_tree = rpart(Sales ~ .,data = Carseats[train,],
                 method="class")

sales_tree
```
***
- We can visualize our tree model with `rpart.plot()`

```{r}
rpart.plot(sales_tree)
```

***
- Behind the scenes `rpart()` is automatically applying a range of cost complexity $\alpha$  values to prune the tree. 

- To compare the error for each  $\alpha$  value, it performs a 10-fold CV (by default)

```{r}
sales_tree$cptable
```

- Here we don't find much improvement after 3 terminal nodes 


***

- Thus, we could use a tree with just 3 terminal nodes and reasonably expect to experience similar results within a small margin of error.


```{r}
plotcp(sales_tree)
```

***
- Grow another tree with `maxdepth=2`

```{r}
sales_tree2 = rpart(Sales ~ .,data = Carseats[train,],
                 method="class", maxdepth=2)

sales_tree2
```
***
- We can visualize our tree model with `rpart.plot()`

```{r}
rpart.plot(sales_tree2)
```


***
- Make predictions on the test data using both trees

```{r}
sales_pred <- data.frame("p_hat"=predict(sales_tree,
                Carseats[-train,],type = "prob")[,"High"],
                        "predicted"=predict(sales_tree, 
                Carseats[-train,], type = "class"),
                        "actual"=Carseats[-train,"Sales"])

sales_pred2 <- data.frame("p_hat"=predict(sales_tree2,
                Carseats[-train,],type = "prob")[,"High"],
                      "predicted"=predict(sales_tree2, 
                Carseats[-train,], type = "class"),
                      "actual"=Carseats[-train,"Sales"])
```

***
- Call the `confusion matrix`

```{r}
cm <- confusionMatrix(data=sales_pred$predicted,
                reference=sales_pred$actual,
                positive="High")
cm$table
```

***
- Call the `confusion matrix`

```{r}
cm2 <- confusionMatrix(data=sales_pred2$predicted,
                reference=sales_pred2$actual,
                positive="High")
cm2$table
```

***
- Performance metrics
```{r}
c(cm$overall[1],cm$byClass[c(1,2,7)])
c(cm2$overall[1],cm2$byClass[c(1,2,7)])
```

***
- ROC curve

```{r echo=FALSE}
pred = prediction(sales_pred$p_hat, sales_pred$actual,
                  label.ordering = c("Low","High"))
roc = performance(pred,"tpr","fpr")
plot(roc, colorize = T, lwd = 2)
abline(a = 0, b = 1)
auc = performance(pred, measure = "auc")
subtitle = sprintf("AUC: %f", auc@y.values)
mtext(side=3,line=1,at=0,adj=0,cex=0.7,subtitle)
```

***
- ROC curve

```{r echo=FALSE}
pred = prediction(sales_pred2$p_hat, sales_pred2$actual,
                  label.ordering = c("Low","High"))
roc = performance(pred,"tpr","fpr")
plot(roc, colorize = T, lwd = 2)
abline(a = 0, b = 1)
auc = performance(pred, measure = "auc")
subtitle = sprintf("AUC: %f", auc@y.values)
mtext(side=3,line=1,at=0,adj=0,cex=0.7,subtitle)
```