---
title: "NBA 4920/6921 Lecture 15"
subtitle: "Elastic Net Application" 
author: "Murat Unal"
date: "10/21/2021"
output: 
  beamer_presentation:
    colortheme: beaver
    df_print: kable
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = '', 
                      warning = FALSE, 
                      message = FALSE)
```

------------------------------------------------------------------------

```{r message=FALSE, warning=FALSE}
rm(list=ls())
options(digits = 3, scipen = 999)
library(tidyverse)
library(ISLR)
library(ggplot2)
library(jtools)
library(caret)
library(leaps)
library(glmnet)
Hitters <- ISLR::Hitters
Hitters <- na.omit(Hitters)
set.seed(2)
```

```{r}
x=model.matrix(Salary~.,Hitters)[,-1]
y=Hitters$Salary
```

```{r}
train=sample(1:nrow(x), 0.7*nrow(x))
```

***
- Previous test RMSEs:

```{r}
RMSE <- matrix(NA,ncol = 1, nrow = 8)
rownames(RMSE) <- c("rmse.ridge.lambdabig",
"rmse.ridge.lambda4","rmse.ridge.lambda0",
"rmse.ridge.lambdabest","rmse.lasso.lambda1se",
"rmse.lasso.lambdabest", "rmse.elnet.lambda1se",
"rmse.elnet.lambdabest")
RMSE[1:6,1] <- c(405,296,300,292,334,297)
RMSE
```


# Elastic net

- Now, there are two parameters to tune: $\lambda$ and $\alpha$.

- The __glmnet__ package allows to tune $\lambda$ via cross-validation for a fixed $\alpha$, but it does not support $\alpha$-tuning.


***
- Let's write our own loop that does the tuning

- First, we create a common `fold_id`, which just allows us to apply the same CV folds to each model.

- We then create a tuning grid that searches across a range of $\alpha$s from 0-1, and empty columns where we’ll dump our model results into.

```{r}
# maintain the same folds across all models
fold_id <- sample(1:10, size = length(y[train]), 
                                    replace=TRUE)

# search across a range of alphas
tuning_grid <- data.frame(
  alpha      = seq(0, 1, by = .1),
  mse_min    = NA,mse_1se    = NA,
  lambda_min = NA,lambda_1se = NA)
```

***
- Now we can iterate over each $\alpha$ value, apply a CV elastic net, and extract the minimum and one standard error MSE values and their respective 
 $\lambda$ values.

```{r}
for(i in seq_along(tuning_grid$alpha)  ) {
  # fit CV model for each alpha value
  fit <- cv.glmnet(x[train,], y[train], 
                   alpha = tuning_grid$alpha[i], 
                                foldid = fold_id)
  
  # extract MSE and lambda values
  tuning_grid$mse_min[i]    <- fit$cvm[fit$lambda==
                                         fit$lambda.min]
  tuning_grid$mse_1se[i]    <- fit$cvm[fit$lambda==
                                         fit$lambda.1se]
  tuning_grid$lambda_min[i] <- fit$lambda.min
  tuning_grid$lambda_1se[i] <- fit$lambda.1se
}

```
***
```{r}
tuning_grid %>% arrange(mse_min)
```

***
- Extract the optimum $alpha$ and $\lambda$ values

```{r}
best.index <- which.min(tuning_grid$mse_min)
best.alpha <- tuning_grid[best.index ,"alpha"]
best.lambda <- tuning_grid[best.index ,"lambda_min"]
best.lambda.1se <- tuning_grid[best.index ,"lambda_1se"]

best.alpha
best.lambda
best.lambda.1se 
```
***
- Now that have identified the preferred model, we retrain the model and simply use `predict` to predict the same model on a new data set.

```{r}
elnet.mod <- glmnet(x[train,], y[train],alpha=best.alpha)

elnet.pred.lambdabest <- predict(elnet.mod, 
                    s=best.lambda,newx=x[-train,])

elnet.pred.lambda1se <- predict(elnet.mod, 
                    s=best.lambda.1se,newx=x[-train,])

rmse.elnet.lambdabest<-sqrt(mean((y[-train] - 
                              elnet.pred.lambdabest)^2))
rmse.elnet.lambda1se<-sqrt(mean((y[-train] -
                              elnet.pred.lambda1se)^2))
```

***
```{r}
RMSE[7:8,1]<-c(rmse.elnet.lambda1se,
               rmse.elnet.lambdabest)
RMSE
```

***
- We could also use the __caret__ package to do cross-validation for both $\alpha$ and $\lambda$

- The package has the `train()` meta engine (aggregator) that allows us to apply almost any direct engine with `method()`

***
```{r}
cv_10 = trainControl(method = "cv", number = 10)

grid =  expand.grid(alpha = seq(0,1,by=0.1),
                    lambda = 10^seq(3,-2,length=100))
elnet = train(
  Salary ~ .,
  data = Hitters[train,],
  method = "glmnet",
  trControl = cv_10,
  preProcess = c("center", "scale"),
  tuneGrid = grid)
```

***
```{r}
elnet$bestTune
alpha.best <- unlist(unname(elnet$bestTune[1]))
lambda.best <- unlist(unname(elnet$bestTune[2]))
```
***
```{r}
plot(elnet, xvar = "lambda")
```

***
- Final model with cross-validated parameters

```{r}
elnet.final <- glmnet(x[train,],y[train],alpha=alpha.best)

elnet.final.lambdabest <- predict(elnet.final, 
                    s=lambda.best,newx=x[-train,])

elnet.lambdabest.caret <- sqrt(mean((y[-train]-
                        elnet.final.lambdabest)^2 ))
elnet.lambdabest.caret
```

***
```{r}
predict(elnet.final,s=lambda.best,type = "coefficients")[1:20,]
```










